---
output:
  github_document:
    html_preview: FALSE
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
```
# citesdb

```{r authors, echo = FALSE, results = 'asis'}
unclass(desc::desc_get_authors(here::here("DESCRIPTION"))) %>% 
  purrr::keep(~"aut" %in% .$role) %>% 
  purrr::map_chr(~paste(.$given, .$family)) %>% 
  glue::glue_collapse(sep = ", ", last = ", and ") %>% 
  paste0("Authors: _", ., "_") %>% 
  cat()
```

[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg){data-external="1"}](https://opensource.org/licenses/MIT)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg){data-external="1"}](http://www.repostatus.org/#wip)
[![CircleCI](https://circleci.com/gh/ecohealthalliance/citesdb.svg){data-external="1"}](https://circleci.com/gh/ecohealthalliance/citesdb)
  [![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ecohealthalliance/citesdb?branch=master&svg=true){data-external="1"}](https://ci.appveyor.com/project/ecohealthalliance/citesdb)
[![codecov](https://codecov.io/gh/ecohealthalliance/citesdb/branch/master/graph/badge.svg){data-external="1"}](https://codecov.io/gh/ecohealthalliance/citesdb)
[![CRAN status](https://www.r-pkg.org/badges/version/citesdb){data-external="1"}](https://cran.r-project.org/package=citesdb)

**citesdb** provides access to the full history of CITES shipments for analysis.


Features:

-  Local caching of database
-  RStudio connection pane with browsable database
-  HTMLwidget-enhanced help files for browsing metadata

While this repository is private you must have a GitHub Personal Access Token
(`GITHUB_PAT`) installed in order to access data.  Instructions for setting this up are found
[here](https://happygitwithr.com/github-pat.html).


## Installation

Install the **citesdb** package with this command:

```{r install_me, eval = FALSE}
source("https://install-github.me/ecohealthalliance/citesdb")
```

```{r message=FALSE, warning=FALSE, error=FALSE, include=FALSE}
options(width=120)
```

## Getting the data

When you first load the package you will see a message like this:

    library(citesdb)
    #> Local CITES database empty or corrupt. Download with cites_db_download()

Not to worry, just do as it says and run `cites_db_download()`.  This will
fetch the most recent database from online.  Currently this is test data that
is approximately 80MB.  It will expand to over 400MB in the database. During
up to 1GB of disk space may be used temporarily.

Once you fetch the data you can connect to the database with the `cites_db()`
command.  If you have the **dplyr** package installed, you can use the
`cites_shipments()` command to load a remote `tibble` that is backed by the
database but not loaded into R.  You can use this
to analyze CITES data without ever loading it into memory, then gather your
results with `collect()`:

```{r, warning = FALSE}
library(citesdb)
library(dplyr)
start <- Sys.time()

cites_shipments() %>%
  group_by(year) %>%
  summarize(number = n()) %>%
  arrange(desc(year)) %>% 
  collect()

stop <- Sys.time()
```

The back-end database, [MonetDB](https://monetdb.org) is very fast and powerful, making such
analyses quite snappy even on such large data and normal desktops and laptops. Here's the
timing of the above query on an ordinary laptop:

```{r}
stop - start
```

If you are using RStudio interactively, loading the CITES package also brings
up a browable pane in the "Connections" tab that lets you explore and preview
the database, as well.

### Contributing

Want have feedback or want to contribute? Great! Please take a look at the [contributing guidelines](https://github.com/ecohealthalliance/citesdb/blob/master/.github/CONTRIBUTING.md) before filing an issue or pull request. 

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/ecohealthalliance/citesdb/blob/master/.github/CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.
