---
output:
  github_document:
    html_preview: FALSE
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
```
# citesdb

```{r authors, echo = FALSE, results = 'asis'}
unclass(desc::desc_get_authors(here::here("DESCRIPTION"))) %>% 
  purrr::keep(~"aut" %in% .$role) %>% 
  purrr::map_chr(~paste(.$given, .$family)) %>% 
  glue::glue_collapse(sep = ", ", last = ", and ") %>% 
  paste0("Authors: _", ., "_") %>% 
  cat()
```

[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg){data-external="1"}](https://opensource.org/licenses/MIT)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg){data-external="1"}](http://www.repostatus.org/#wip)
[![CircleCI](https://circleci.com/gh/ecohealthalliance/citesdb.svg){data-external="1"}](https://circleci.com/gh/ecohealthalliance/citesdb)
  [![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ecohealthalliance/citesdb?branch=master&svg=true){data-external="1"}](https://ci.appveyor.com/project/ecohealthalliance/citesdb)
[![codecov](https://codecov.io/gh/ecohealthalliance/citesdb/branch/master/graph/badge.svg){data-external="1"}](https://codecov.io/gh/ecohealthalliance/citesdb)
[![CRAN status](https://www.r-pkg.org/badges/version/citesdb){data-external="1"}](https://cran.r-project.org/package=citesdb)


Features:
-  Local caching of versions of database
-  Metadata in **[dataspice](https://github.com/ropenscilabs/dataspice)**/<schema.org> format
-  RStudio connection pane with browsable database
-  HTMLwidget-enhanced help files for browsing metadata?


## Installation

Install the **citesdb** package with this command:

```{r install_me, eval = FALSE}
source("https://install-github.me/ecohealthalliance/citesdb")
```

```{r message=FALSE, warning=FALSE, error=FALSE, include=FALSE}
options(width=120)
```

## Usage

```{r example}
#library(citesdb)

# some example code
```


# Current Verison

```{r version}
#packageVersion("citesdb")
```

## Contributing

Want have feedback or want to contribute? Great! Please take a look at the [contributing guidelines](https://github.com/ecohealthalliance/citesdb/blob/master/.github/CONTRIBUTING.md) before filing an issue or pull request. 

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/ecohealthalliance/citesdb/blob/master/.github/CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.
