---
title: "Exploring CITES Trade Data with citesdb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring CITES Trade Data with citesdb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**citesdb** is a package to conveniently analyze the full CITES shipment-level wildlife trade database, available at <https://trade.cites.org/>. This data consists of over 20 million records of reported shipments of wildlife and wildlife
products subject to oversight under the [Convention on International Trade in Endangered Species of Wild Fauna and Flora](https://www.cites.org).

## Installation

Install the **citesdb** package with this command:

```{r install_me, eval = FALSE}
source("https://install-github.me/ecohealthalliance/citesdb")
```

```{r message=FALSE, warning=FALSE, error=FALSE, include=FALSE}
options(width=120)
knitr::opts_chunk$set(cache = TRUE)
citesdb::cites_disconnect()
MonetDBLite::monetdblite_shutdown()
```

## Usage

### Getting the data

When you first load the package you will see a message like this:

    library(citesdb)
    #> Local CITES database empty or corrupt. Download with cites_db_download()

Not to worry, just do as it says and run `cites_db_download()`.  This will
fetch the most recent database from online, an approximately 158 MB download.
It will expand to over 1 GB in the local database. During
the download and database building up to 3.5 GB of disk space may be used temporarily.

### Using the database

Once you fetch the data you can connect to the database with the `cites_db()`
command.  You can use the
`cites_shipments()` command to load a remote `tibble` that is backed by the
database but not loaded into R.  You can use this
to analyze CITES data without ever loading it into memory, then gather your
results with `collect()`. For example:

```{r getdata, include = FALSE}
if (!citesdb::cites_status()) citesdb::cites_db_download()
```

```{r, warning = FALSE}
library(citesdb)
library(dplyr)
start <- Sys.time()

cites_shipments() %>%
  group_by(Year) %>%
  summarize(number = n()) %>%
  arrange(desc(Year)) %>%
  collect()

stop <- Sys.time()
```

(_Note that running `collect()` on all of `cites_shipments()` will load a >3GB data frame into memory!_)

The back-end database, [MonetDB](https://monetdb.org), is very fast and powerful, making such
analyses quite snappy even on such large data using normal desktops and laptops. Here's the
timing of the above query, which processes over 20 million records, on an ordinary laptop:

```{r}
stop - start
```

If you are using a recent version of RStudio interactively, loading the CITES package also brings
up a browsable pane in the "Connections" tab that lets you explore and preview
the database, as well as interact with it directly via SQL commands.

If you don't need any of the bells and whistles of this package, you can download
the raw data as a single compressed TSV from the [releases page](https://github.com/ecohealthalliance/citesdb/releases),
or as a `.zip` file of many CSV files from original source at <https://trade.cites.org/>.

### Metadata

The database also contains tables of field metadata, codes used, and CITES countries. This information
comes from ["A guide to using the CITES Trade Database"](https://trade.cites.org/cites_trade_guidelines/en-CITES_Trade_Database_Guide.pdf),
on the CITES website.  Convenience functions `cites_metadata()`, `cites_codes()`, and `cites_parties()`
access this information:

```{r}
head(cites_metadata())

head(cites_codes())
```

More information on the release of shipment-level data can be found in the
`?guidance` help file.


## Pitfalls

The CITES shipment database is a valuable resource for understanding part of
the global wildlife trade.  However, care must be taken in working with and
interpreting this data.

### Repeated records of the same shipment

Single shipments may be recorded multiple times, as they may be reported by
both exporting and importing countries.  Take, for example, this set of
shipments of orchids from a U.S. exporter to the Netherlands:

```{r}
cites_shipments() %>% 
  filter(Year == 2016, Export.permit.RandomID == "ce63001ff5", Genus == "Paphiopedilum") %>% 
  collect() %>% 
  head() %>% 
  knitr::kable()
```

See that each pair of shipments is identical, except that the `Reporter.type`,
indicating that the shipment was reported both from the U.S. and from the
Netherlands.  Note that while the importer reported both import and export
permit numbers, the exporter reported only the export permit numbers.

How to deal with repeat records? If you are aggregating records to calculate
total shipments it is best to remove them, but not every record is recorded twice.
This code filters to use only the importer's records, should all fields other
than `Reporter.type` and `Import.permit.RandomID` be the same:

```{r}
cites_shipments() %>% 
  filter(Year == 2016, Export.permit.RandomID == "ce63001ff5", Genus == "Paphiopedilum") %>% 
  collect() %>% 
  group_by_at(vars(-Reporter.type, -Import.permit.RandomID)) %>% 
  mutate(n = n()) %>% 
  filter((n > 1 & Reporter.type == "I") | n == 1) %>% 
  ungroup() %>% 
  head() %>% 
  knitr::kable()
```

## Related work

The [**rcites**](https://github.com/ropensci/rcites) package provides
access to the Speciesplus/CITES Checklist API, which includes metadata about
species and their protected status through time.

## Citation

If you use **citesdb** in a publication, please cite both the package and source data:

```{r, results = "asis", echo = FALSE}
print(citation("citesdb"), style = "textVersion")
```

```{r message=FALSE, warning=FALSE, error=FALSE, include=FALSE}
cites_disconnect()
MonetDBLite::monetdblite_shutdown()
```
